### 2023-12-26 misc/expo/additive.tex
Furstenberg--Sarkozy lemma

>>>>

\advsect The Furstenberg--S\'ark\"ozy theorem

In this section, we'll prove a bound on the size of {\it square-difference-free sets}. These are sets
$A\subseteq \NN$ such that $A-A$ does not contain any of the squares $\{1,4,9,16,\ldots\}$.
The proof, which is due to B.~Green, T.~Tao, and T.~Ziegler,
makes use of a density increment argument similar to the one used in the proof of
Roth's theorem, but does not use any Fourier analysis. We begin with the following lemma, which allows
us to increment the density of a set containing a square in its difference set.

\proclaim Lemma~\advthm. There exists a real constant $c>0$ such that the following holds.
Suppose that there is $N\in \NN$ and $A\subseteq [N]$ of density $\delta = |A|/N$ such that
$A-A$ does not contain any positive square numbers. Then there exists an arithmetic progression
$P$ of length $\lfloor N^{1/3}\rfloor$ such that $|A\cap P| \ge (\delta + c\delta^3)|P|$.

Suppose that for
some $\delta_0>0$ such that for arbitrarily large $N\in \NN$, we can find $A\subseteq [N]$
of size $|A|\ge \delta_0 N$ such that $A-A$ does not contain a square. Then for
arbitrarily large $N'\in \NN$, there exists a subset
$A'\subseteq [N']$ such that $|A'|\ge (\delta_0 + c{\delta_0}^3)N'$ and $A'-A'$ does not contain a square.

\proof Suppose $\delta_0>0$ is such that for arbitrarily large $N$ we can find $A\subseteq [N]$
of density $\delta = |A|/N \ge \delta_0$ such that $A-A$ does not contain a square. Then for all
$n\in [N]$, we have
$$ \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]} \one_A(n) \one_A\bigl( n+(r+h)^2\bigr) = 0.$$
Now for $n\in [N]$, note that $n+(r+h)^2$ is in $[N]$ for all
$n\le N-(N^{1/3}-N^{1/100})^2 = N-O(N^{2/3})$. This means that
$$ \ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}
\one_{[N]}\bigl( n+(r+h)^2\bigr) = 1 - O(N^{-1/3}).$$
By further requiring that $n\in A$ inside the expectation, we have
$$ \ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}
\one_A(n)\one_{[N]}\bigl( n+(r+h)^2\bigr) = \delta - O(N^{-1/3}),$$
since the proportion of $n\in [N]$ that are also in $A$ is $\delta$.
Lastly, by a simple change of variables we have
$$\eqalign{
\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} &\ex_{h\in [N^{1/100}]} \one_A\bigl(n+(r+h)^2\bigr) \cr
&=\ex_{m\in A} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]} \one_{[N]}\bigl(m-(r+h)^2\bigr) \cr
&= \delta - O(N^{-1/3}). \cr
}$$
Set $f = \one_A - \delta\one_{[N]}$. From the above computations we deduce that
$$\eqalign{
\ex_{n\in [N]} &\ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}
f(n) f\bigl( n+(r+h)^2\bigr) \cr
&= 
\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}\cr
&\qquad\qquad\Bigl(\one_A(n)\one_A\bigl(n+(r+h)^2\bigr)
-\delta\one_A(n)\one_{[N]}\bigl(n+(r+h)^2\bigr) \cr
&\qquad\qquad\qquad\qquad\qquad- \delta\one_A\bigl(n+(r+h)^2\bigr)
+ \delta^2\one\bigl(n+(r+h)^2\bigr)\Bigr) \cr
&= 0 - \delta^2 - \delta^2 +\delta^2 + O(N^{-1/3}) \cr
&= -\delta^2 + O(N^{-1/3}) .\cr
}$$
The triangle inequality then gives
$$ \ex_{n\in [N]} \bigl| f(n)\bigr| \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in [N^{1/100}]} f\bigl( n+(r+h)^2\bigr)
\Bigr| \ge \delta^2-O(N^{-1/3}) \gg \delta^2,$$
where here and in the rest of the proof, the Vinogradov notation hides an absolute constant that works
for all $N\in \NN$. We also have
$$\eqalign{
\ex_{n\in [N]} \bigl| f(n)\bigr| ^2
&= \ex_{n\in [N]} \Bigl(\one_A(n) - 2\delta \one_A(n)\one_{[N]}(n) + \delta^2 \one_{[N]}(n)\Bigr)\cr
&= \delta-\delta^2, \cr
}$$
so by the Cauchy--Schwarz inequality,
$$\eqalign{
\delta^4-O(N^{1/3}) &\le
\biggl(\ex_{n\in [N]} \bigl| f(n)\bigr| \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in [N^{1/100}]} f\bigl( n+(r+h)^2\bigr)
\Bigr|\biggr)^2 \cr
&\le \ex_{n\in [N]} \bigl| f(n) \bigr|^2 
\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in N^{1/100}} f\bigl(n+(r+h)^2\bigr)\Bigr|^2\cr
&= (\delta-\delta^2)
\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in N^{1/100}} f\bigl(n+(r+h)^2\bigr)\Bigr|^2,\cr
}$$
and therefore
$$
\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in N^{1/100}} f\bigl(n+(r+h)^2\bigr)\Bigr|^2
\gg {\delta^4\over \delta-\delta^2} \ge \delta^3.$$
We can rewrite this as
$$
\ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} 
f\bigl(n+(r+h)^2\bigr) f\bigl(n+(r+h')^2\bigr) \gg \delta^3.
$$
Replacing $n$ by $n-(r+h)^2$ in the innermost expectation introduces an error term of at most $O(N^{-1/3})$
and deleting the diagonal case $h=h'$ introduces an error term of at most $O(N^{-1/100})$,
so we have
$$\eqalign{
\ex_{h,h'\in N^{1/100}}&\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} 
f\bigl(n+(r+h)^2\bigr) f\bigl(n+(r+h')^2\bigr) \cr
&\le
\ex_{h,h'\in N^{1/100}}\Bigl|\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} 
f\bigl(n+(r+h)^2\bigr) f\bigl(n+(r+h')^2\bigr) \Bigr|\cr
&\le
\ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
&\qquad\qquad f(n) f\bigl(n-(r+h)^2+(r+h')^2\bigr) + O(N^{-1/3})\cr
&= \ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/3})\cr
&= \ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/3})\cr
&= \ex_{h\ne h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} \cr
&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/100}),\cr
}$$
whence
$$\ex_{h\ne h'\in N^{1/100}}\Bigl|\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}
f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) \Bigr| \gg \delta^3 $$
by the triangle inequality. From this we see that
there are distinct $h$ and $h'$ in $[N^{1/100}]$ with
$$\Bigl|\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}
f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr)\Bigr| \gg \delta^3$$
Here another application of the triangle inequality and the fact that $\bigl|f(n)\bigr| \le 1$ together
give
$$\ex_{n\in [N]} \Bigl|\ex_{r\in [N^{1/3}]}
f\bigl(n- (h'-h)(2r+h'-h)\bigr)\Bigr|\gg \delta^3.$$
Now we will perform another shift on $n$, this time by $(h'-h)(h'+h) = O(N^{1/50})$. In the expectation
over $n$, this produces an error term of at most $O(N^{-49/50})$, which we can hide behind the Vinogradov notation
as well. Letting $d = 2(h'-h)$, we have
$$\eqalign{
\ex_{n\in [N]} \Bigl| &\ex_{r\in [N^{1/3}]} f(n+dr)\Bigr| \cr
&= \ex_{n\in [N]} \Bigl| \ex_{r\in [N^{1/3}]} f\bigl(n+2(h'-h)r\bigr)\Bigr| \cr
&\ge \ex_{n\in [N]} \Bigl| \ex_{r\in [N^{1/3}]} f\bigl(n+(h'-h)(2r+h+h')\bigr)\Bigr| - O(N^{-49/50}) \cr
&\gg \delta^3 \cr
}$$
But recall that we have
$$\sum_{n\in [N]} f(n) = 0,$$
so since $d \le 2N^{1/100}$ and $r\le N^{1/3}$, we have
$$\sum_{n\in [N]} f(n+dr) = O(N^{1/3+ 1/100})$$
for all $r\in [N^{1/3}]$, and hence
$$\ex_{n\in [N]} \ex_{f\in [N^{1/3}]} f(n+dr) = O(N^{-2/3+ 1/100}).$$
Then since $\bigl(x+|x|\bigr)/2 = \max(x,0)$ for all $x\in \RR$, we have
$$ \eqalign{
\ex_{n\in [N]} \max\Bigl( \ex_{r\in [N^{1/3}]} &f(n+dr), 0\Bigr) \cr
&\gg {\delta^3 + O(N^{-2/3+1/100})\over 2} \cr
&\gg \delta^3.\cr
}$$
By artificially zeroing out all terms corresponding to the largest $2N^{1/3+1/100}$ elements of the interval $[N]$,
we introduce an error
term of $N^{-2/3+1/100}$, which disappears behind the asymptotic notation.
Then by the pigeonhole principle we may find $n\in [N-2N^{1/3+1/100}]$ such that
$$\ex_{r\in [N^{1/3}]} f(n+dr) \gg \delta^3.$$
Let $P\subseteq [N]$ be the arithmetic progression $\bigl\{n + dr : r\in [N^{1/3}]\bigr\}$ and let
$c$ be the absolute constant implied by the Vinogradov notation. We have
$$\ex_{x\in P} f(x) \ge c\delta ^3,$$
and can unravel the definition of $f$ to obtain
$${|A\cap P|\over |P|} \ge \delta + c\delta^3,$$
which is what we wanted.\slug

### 2023-12-26 misc/expo/entropy.tex
Furstenberg--Sarkozy lemma

\maketitle{Entropy and the polynomial Freiman--Ruzsa theorem}{by}{Marcel K. Goh}{23 November 2023}

\floattext4.5 \ninebf Note.
\ninepoint This is a gentle introduction to the notion of entropy as it is used
in combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. The preliminary portion of
these notes are largely transcribed from lectures given by W.~T.~Gowers.

@@

Immediate from the definition of the conditional entropy $\Eta\{Y\given X\}$ given
in the additivity axiom is that if
$X$ and $Y$ are independent, then $\Eta\{Y\given X\} = \Eta\{Y\}$. But since the uniform distribution
on $\{0,1\}^n$ is the joint distribution $(X_1,\ldots,X_n)$ where the $X_i$ are independent and
uniformly distributed on $\{0,1\}$, we can additionally use the normalisation axiom to conclude that
$$\Eta\{X\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
by induction. Induction also yields the chain rule
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\} + \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.$$

The first proposition establishes the intuitive fact
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

@@

\proof

>>>>

\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{23 November 2023}

\floattext4.5 \ninebf Note.
\ninepoint This is a gentle introduction to the notion of entropy as it is used
in additive combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. The preliminary portion of
these notes are largely transcribed from lectures given by W.~T.~Gowers.

@@

Notationally, we would expect that $\Eta\{Y\given X\} = \Eta\{Y\}$ if $X$ and $Y$ are independent.
This is the first proposition we will carefully prove using the axioms above.

\proclaim Let $X$ and $Y$ be independent random variables. Then
$\Eta\{Y\given X\} = \Eta\{Y\}$ and consequently
$\Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$.

\proof Suppose $X$ takes values in a finite set $A$. Then for all $x\in A$, the distribution of
$Y$ and $Y$ given that $X = x$ is the same, so
$$\Eta\{Y \given X\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\given X = x\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\}
=\Eta\{Y\}.$$
The second version of the statement is a consequence of the additivity axiom.\slug

We will sometimes use the notation $X^n$ to denote the vector $(X_1,\ldots,X_n)$ where the $X_i$
are independent copies of the random variable $X$.

\proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.

\proof Perform induction on $n$ and use the previous proposition.\slug

In the case where $X$ is the uniform distribution on $\{0,1\}$, and $X_1,\ldots,X_n$ are independent
and distributed as $X$,
we can additionally use the normalisation axiom to conclude that
$$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
by induction. For general random variables $X_1,\ldots,X_n$, induction also yields the {\it chain rule}
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\} + \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.$$

Next we show that a constant random variable has zero entropy. This reflects the idea that we get
no information from sampling a random variable if it always takes the same value no matter what.

\proclaim Proposition \advthm. Let $X$ be a discrete random variable.
Then $\Eta\{X\} = 0$ if and only if it takes exactly one value.

\proof First suppose that $X$ takes only one value.
Let $a$ be the value of $X$ such that $\pr\{X=a\} = 1$.
Then $(X,X)$ equals $(a,a)$ with probability $1$ as well,
so $\Eta\{X\} = \Eta\{X,X\}$ by the invariance axiom. But it can easily be checked that
$X$ and $(X,X)$ are independent (we have, for example, $\pr\bigl\{X = a, (X,X)= (a,a)\bigr\} =
\pr\{X = a\}\pr\bigl\{(X,X) = (a,a)\bigr\}$), so $\Eta\{X,X\} = 2\Eta\{X\}$. Thus we conclude
that $\Eta\{X\} = 0$.

Now suppose that $X$ takes more than one value; let $A$ be the set of $a$ such that
$\pr\{X = a\} > 0$.
\slug

Now we establish the intuitive fact
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

@@

\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
taking values in $B$. Then by the maximisation axiom, $\eta\{X\}\le \eta\{Y\}$, since $Y$ is
uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
are the same random variable.
Now say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
we have $\Eta\{X\} = 0$, and $Y$ takes at least two values,
